\documentclass[../main-v1.tex]{subfiles}
\begin{document}
\chapter{Global computing strategy}

\hideme{I propose moving the subsections here into the appropriate later chapters and using this section to describe the global compute model}

\dword{dune} is a global collaboration with contributions from institutions worldwide.  The large scale computing effort relies on multi-national contributions. The overall strategy for computing resources is for primary raw data storage to reside at the large host labs (CERN and FNAL) with computing resources such as disk and CPU largely contributed by collaborating countries.   CPU contributions are provided by a large number of sites worldwide while storage is concentrated at a few larger international sites. Table \ref{tab:coop:disk} shows the distribution of disk space pledges for 2021 and 2022 while Tables \ref{tab:coop:sites} and \ref{tab:coop:ussites}shows the sites contributing CPU resources. 

The large LHC experiments have historically relied on a tiered structure, with national Tier-1 centers serving regional Tier-2 and Tier-3 centers.  The DUNE model builds on the emergence of faster networks to move to  a service-oriented model, where sites provide services, disk, CPU, real memory/core and archival tape, and projects are  distributed to them based on their capabilities and available networking.  For example, a site with large CPU and memory/core but slower networking would be ideal for simulation while small memory/core and fast access to large local disk stores would be ideal for high level data analysis.  This model requires a high level view of data locations and job placement with continual monitoring for bottlenecks but allows new sites to contribute in an optimal way. 

This model builds on the highly successful tools developed by the WLCG and OSG for LHC computing.  As DUNE's needs are substantially smaller that those of the large LHC experiments we can be confident that the existing infrastructure can be incrementally augmented to meet DUNE's requirements. 


\begin{dunetable}
[National Disk pledges]{llrr}{tab:coop:disk}
{Disk pledges in PB for 2021 and 2022 .}
Country/Lab	&	Name	&	2021	&	2022	\\
FNAL	&	FNAL	&	2.2	&	7.6	\\
CERN	&	CCERN	&	2.2	&	1.6	\\
BNL	&	BNL	&	0.5	&	0.5	\\
UK	&	GridPP	&	4.0	&	4.0	\\
FR	&	CC-IN2P3	&	0.5	&	0.5	\\
ES	&	PIC Tier-1	&	0.5	&	0.72	\\
NL	&	NL/LHC Tier-1	&	1.9	&	1.8	\\
CZ	&	CZ-Prague-T2	&	0.3	&	1.0	\\
%BR	&	CBPF	&	0	&		\\
IN	&	TIFR	&	0.75&	0.75\\
RU	&	JINR	&	-	&	0.5	\\
\hline
Total pledge	&		&	12.85	&	18.97	\\
\end{dunetable}

%\tiny
\begin{dunetable}
[List of DUNE Compute Sites]
{l l r l}{tab:coop:sites}
{List of non-US international DUNE compute sites as of December 2021.  Sites with substantial rucio controlled disk are noted.}
Site name	&	RC Site	&	Disk	&	Country	\\
\hline
BR\_CBPF	&	BR\_CBPF	&		&	Brazil\\
BR\_UNICAMP	&BR\_UNICAMP	&		&	Brazil\\
CA\_Victoria	&	CA\_Victoria	&		&	Canada\\
CERN	&	CERN-PROD	&	Yes	&	Switzerland\\
CH\_UNIBE-LHEP	&	UNIBE-LHEP	&		&	Switzerland\\
CZ\_FZU	&	FZU	&	Yes	&	Czechia\\
ES\_CIEMAT	&	CIEMAT-LCG2	&		&	Spain\\
ES\_PIC	&	pic	&	Yes	&	Spain\\
FR\_CCIN2P3	&	IN2P3-CC	&	Yes	&	France\\
IN\_TIFR	&	IN\_TIFR	&	Yes	&	India\\
NL\_NIKHEF	&	NIKHEF-ELPROD	&		&	Netherlands\\
NL\_SURFsara	&	SURFsara	&	Yes	&	Netherlands\\
RU\_JINR	&	JINR\_CONDOR\_CE	&	Yes	&	Russian Federation\\
UK\_Bristol	&	UKI-SOUTHGRID-BRIS-HEP	&		&	United Kingdom\\
UK\_Brunel	&	UKI-LT2-Brunel	&		&	United Kingdom\\
UK\_Edinburgh	&	UKI-SCOTGRID-ECDF	&		&	United Kingdom\\
UK\_Imperial	&	UKI-LT2-IC-HEP	&		&	United Kingdom\\
UK\_Lancaster	&	UKI-NORTHGRID-LANCS-HEP	&	Yes	&	United Kingdom\\
UK\_Liverpool	&	UKI-NORTHGRID-LIV-HEP	&		&	United Kingdom\\
UK\_Manchester	&	UKI-NORTHGRID-MAN-HEP	&	Yes	&	United Kingdom\\
UK\_Oxford	&	UKI-SOUTHGRID-OX-HEP	&		&	United Kingdom\\
UK\_QMUL	&	UKI-LT2-QMUL	&	Yes	&	United Kingdom\\
UK\_RAL-PPD	&	UKI-SOUTHGRID-RALPP	&		&	United Kingdom\\
UK\_RAL-Tier1	&	RAL-LCG2	&	Yes	&	United Kingdom\\
UK\_Sheffield	&	UKI-NORTHGRID-SHEF-HEP	&		&	United Kingdom\\

\end{dunetable}
\begin{dunetable}
[List of US DUNE Compute Sites]
{l l r l}{tab:coop:ussites}
{List of US DUNE compute sites as of December 2021.  Sites with substantial rucio controlled disk are noted.}
US\_UConn-HPC	&	UConn-HPC	&		&	United States\\
US\_BNL	&	BNL-SDCC-CE01	&	Yes	&	United States\\
US\_Caltech	&	CIT\_CMS\_T2	&		&	United States\\
US\_Clemson	&	Clemson-Palmetto	&		&	United States\\
US\_Colorado	&	UColorado\_HEP	&		&	United States\\
US\_Florida	&	UFlorida-HPC	&		&	United States\\
US\_FNAL	&	GPGrid	&	Yes	&	United States\\
US\_KSU	&	BEOCAT-SLATE	&		&	United States\\
US\_Lincoln	&	Rhino	&		&	United States\\
US\_Michigan	&	AGLT2	&		&	United States\\
US\_MIT	&	MIT\_CMS	&		&	United States\\
US\_MWT2	&	MWT2	&		&	United States\\
US\_Nebraska	&	Nebraska	&		&	United States\\
US\_NERSC & NERSC &   & United States\\
US\_NMSU-DISCOVERY	&	SLATE\_US\_NMSU\_DISCOVERY	&		&	United States\\
US\_NotreDame	&	NWICG\_NDCMS	&		&	United States\\
US\_Omaha	&	Crane	&		&	United States\\
US\_PuertoRico	&	uprm-cms	&		&	United States\\
US\_SU-ITS	&	SU-ITS-CE2	&		&	United States\\
US\_UChicago	&	MWT2	&		&	United States\\
US\_UCSD	&	UCSDT2	&		&	United States\\
US\_Wisconsin	&	GLOW	&		&	United States\\
US\_WSU	&	WSU - GRID\_ce2	&		&	United States\\
\end{dunetable}

\hideme{the following chapter goes into the relevant subsections} 

\end{document}