% label (always all lower case) is used in chapter files as \dword{label}

% Add a `dune word' or a `dune abbreviation' using these models:

%\newduneword{label }{full term}{description with no period at end}

%\newduneabbrev{label}{abbreviation}{full term}{description with no period at end}

% ADDED TO OVERALL GLOSSARY 26 Feb by Anne; REFRESH glossary.tex 
%\newduneabbrev{corsika}{CORSIKA}{Cosmic Ray SImulations for KAscade}{a program for detailed simulation of extensive air showers initiated by high-energy cosmic-ray particles.}

%\newduneabbrev{scd}{SCD}{Scientific Computing Division}{Fermilab's Scientific Computing Division}

%\newduneabbrev{garsoft}{GArSoft}{Gaseous Argon Software}{A software toolkit similar to LArSoft, but targeted at the gaseous argon time projection chamber and calorimeter of ND-GAr}

%\newduneword{ndgarlite}{ND-GAr-Lite}{a temporary muon spetrometer consisting of the magnet and steel flux return of ND-GAr, but with a simplified tracking chamber made with scintillating bars}

%\newduneword{github}{GitHub}{a commerical web service providing code version management, storage, and browsing via git}

%\newduneword{git}{git}{a distributed version-control system, commonly used to manage software}

%\newduneword{xrootd}{xrootd}{XRootD is a high performance data system widely used in HEP to store and to distribute data to jobs.  It allows streaming of data.}

%\newduneabbrev{gpuaas}{GPUAAS}{GPU As A Service}{A technique which allows many non-GPU-enabled compute nodes to share a GPU resource by sending work to it over the network and waiting for results to be returned.}
% END ADDED TO OVERALL GLOSSARY 26 Feb by Anne

% ADDED TO OVERALL GLOSSARY; REFRESH glossary.tex \newduneabbrev{cvmfs}{{\tt cvmfs}}{CernVM File System}{The CernVM File System provides a scalable, reliable and low-maintenance software distribution service. It was developed to assist High Energy Physics (HEP) collaborations to deploy software on the worldwide-distributed computing infrastructure used to run data processing applications} %TRJ put this in an issue; different (shorter) def: A distributed file system designed for scalable, high-performance distribution of software to interactive and batch computers

% ADDED TO OVERALL GLOSSARY; REFRESH glossary.tex\newduneabbrev{mrb}{{\tt mrb}}{Multi Repository Build System}{A Fermilab developed build system based on \dword{cmake} that allows development and builds of code from multiple repositories}
%please don't use \tt

% ADDED TO OVERALL GLOSSARY; REFRESH glossary.tex\newduneabbrev{cmake}{{\tt cmake}}{Cmake}{CMake is an open-source, cross-platform family of tools designed to build, test and package software} %chged to newduneword

%\newduneabbrev{ups}{{\tt ups}}{Unix Product Support}{UPS is a Fermilab developed package manager that supports installation of multiple versions of a product and multiple builds per version.} %TRJ put this in an issue; exists, but should we change definition?

\newduneabbrev{vd}{VD}{Vertical Drift}{Detector using vertical drift TPC technology.}

\newduneabbrev{hd}{HD}{Horizontal Drift}{Detector using horizontal drift TPC technology.}

%\newduneabbrev{hv}{HV}{High Voltage}{High voltage system for TPC or other detector system.}

%\newduneabbrev{cisc}{CISC}{Cryogenic Instrumentation and Slow Controls}{DAQ subgroup responsible for monitoring detector systems. }

\newduneabbrev{poms}{POMS}{Production Operations Management System}{A Fermilab web service interface that enables automated jobs submission on distributed resources and subsequent monitoring and recovery of failed submissions, debugging and record keeping.}

\newduneabbrev{compass}{Compass}{
Common Muon and Proton Apparatus for Structure and Spectroscopy (COMPASS)}{
COMPASS is a multipurpose experiment at CERN’s Super Proton Synchrotron (SPS).}

\newduneabbrev{esnet}{ESnet}{Energy Sciences Network}{The Department of Energy's dedicated science network.}

\newduneword{project.py}{project.py}{xml based job configuration system developed by the MicroBooNE collaboration.}

\newduneabbrev{ppfx}{PPFX}{Package to Predict the FluX}{Fermilab supported package which implements hadron production corrections to geant4 simulations and propagates uncertainties for the NuMI  and LBNF beam lines.}

\newduneabbrev{ml}{ML}{Machine Learning}{Machine Learning}

\newduneword{datalake}{data lake}{data lake}{The concept of a storage service with multiple levels of quality of service where the end user does not have to be aware of where the data is being served from in order to access the data. Data lakes are not a reality yet.}

\newduneabbrev{proto}{ProtoDUNE}{ProtoDUNE}{The protoDUNE detectors at CERN, includes the single phase, dual phase and vertical drift prototypes.} %Anne - not needed; already have {protodune}

\newduneabbrev{metacat}{MetaCat}{MetaCat}{Metadata Catalog, a modern replacement for the file description portion of the sam metadata catalog.}

\newduneword{g4lbnf}{g4lbnf}{LBNF neutrino beamline simulation program\cite{g4lbnf}.}

\newduneword{Geant4Reweight}{Geant4Reweight}{Framework for evaluating and propagating hadronic interaction uncertainties in Geant4\cite{Calcutt:2021zck}.}

\newduneword{geantnet}{G\'EANT}{G\'EANT interconnects Europe's national research and education networking (NREN) organizations with the high bandwidth, high speed and highly resilient pan-European backbone.
}

\newduneabbrev{nren}{NREN}{National Research and Education Network}{National level research computing network infrastructure.}

\newduneabbrev{vrf}{VRF}{Virtual Routing and Forwarding}{Networking overlays that provide  a logical routing infrastructure  that allows flexible traffic engineering.}

\newduneword{samvalue}{value}{A generic quantity describing a file in the \dword{sam} data catalog.}

\newduneword{samparameter}{parameter}{A user or experiment described quantity describing a file in the \dword{sam} data catalog.}

\newduneword{samproject}{SAM-project}{A server process running centrally that maintains a predefined list of files and delivers information about  their locations when asked by distributed processes. The project tracks success and failure of file processing.}

\newduneword{samconsumer}{consumer}{A client process that requests information about file locations from a \dword{samproject}, process the file and reports success or failure.}

\newduneword{samdataset}{SAM-dataset}{ A dynamic collection of files defined by queries to the \dword{sam} data catalog.}
\newduneword{metacatdataset}{MetaCat-dataset}{ A fixed but mutable collection of files defined by queries to the \dword{metacat} data catalog.}

\newduneword{samsnapshot}{SAM-snapshot}{A fixed collection of files corresponding to a \dword{sam} \dword{samdataset} at a particular point in time.}

\newduneabbrev{mql}{MQL}{\dword{metacat} Queryl Language} {A query language which supports queries of the \dword{metacat} data catalog, including parentage and logical functions such as union, join and subtraction.}

\newduneword{pdhd}{ProtoDUNE-HD}{ProtoDUNE with horizontal drift technology.  This refers to the \dword{apa}-based prototype to run in ProtoDUNE Run II.} 

\newduneword{pdvd}{ProtoDUNE-VD}{ProtoDUNE with vertical drift technology.  This refers to the \dword{crp}-based prototype to run in ProtoDUNE Run II.} 

\newduneabbrev{ucondb}{uconDB}{Unstructured Conditions Database}{Unstructured conditions database developed for FNAL fixed target experiements\cite{bib:ucondb}.}

\newduneabbrev{json}{JSON}{JavaScript Object Notation}{Open standard data interchange format that uses pair-value pairs and maps well onto python data formats such as tuples and lists.}

\newduneabbrev{dqmd}{DQMD}{Data Quality and Monitoring Database}{Database storing the results of data quality monitoring.}

\newduneabbrev{db}{DB}{Database}{Database, can have many formats based on optimal use.}

\newduneabbrev{vm}{VM}{Virtual Machine}{Emulator of a physical computer that allows multiple users to configure different operating systems  while sharing physical hardware. }

\newduneword{enstore}{Enstore}{Enstore is a mass storage system developed by Fermilab that provides distributed access and management of data stored on tapes \cite{Litvintsev:2012dw}.}

%\newduneabbrev{nas}{NAS}{Network Attached Storage}{High performance disk available on interactive nodes.}


\newduneword{stashcache}{StashCache}{A distributed caching federation that enables opportunistic users to utilize nearby opportunistic storage\cite{bib:stashcache}}

\newduneabbrev{hdf5}{HDF5}{Hierarchical Data Format}{Data format\cite{bib:hdf5} widely used in \dword{ml}.}

\newduneword{pdcoldbox}{ProtoDUNE-VD coldbox}{  Coldbox used for testing the \dword{pdvd} electronics and \dword{daq} systems.}

\newduneword{protoii}{ProtoDUNE Run II}{2022-23 run of the \dword{proto} prototypes at CERN.}

\newduneword{glideinwms}{GlideinWMS}{GlideninWMS \cite{sfiligoi2009pilot}.}

\newduneword{mars}{MARS}{The MARS code system is a set of Monte Carlo programs for detailed simulation of coupled hadronic and electromagnetic cascades, with heavy ion, muon and neutrino production and interactions.}

\newduneabbrev{fife}{FIFE}{Fabric for Intensity Frontier Experiments}{Fermilab computing infrastructure for Intensity Frontier Experiments \cite{box2014fife}.}

\newduneword{tr}{trigger record}{A data record produced by the \dword{dune} \dword{daq} system.  A Trigger Record can contain multiple interaction "events" or none.}%anne made it lower case

\newduneword{postgres}{Postgres}{PostgreSQL, also known as Postgres, is a free and open-source relational database management system used extensively for databases in HEP.}

\newduneword{hllhc}{HL-LHC}{High-luminosity \gls{lhc}} %anne added

\newduneabbrevs{iov2}{IOV}{interval of validity}{intervals of validity}{Interval over which something is valid} %anne added - plural was wrong - may need to fix orig {iov} too

\newduneword{fhicl}{FHICL}{Fermilab Hierarchical Configuration Language; a standard configuration language for the
storage, communication, and manipulation of scientific parameter sets}  %anne added

\newduneword{hwdb}{HWDB}{hardware database}   %anne added

\newduneword{singlecube}{SingleCube}{add def}  %anne added

\newduneword{s3}{S3}{The Amazon cloud-based commercial storage service.}  %anne added

\newduneword{openstack}{OpenStack}{An open source cloud software used to deploy instances of containers.}  %anne added

\newduneword{cric}{CRIC}{The Computing Resource Information System (CRIC) is a framework providing a centralized (and flexible) way to describe which resources DUNE using and also how DUNE uses them. }  %anne added


\newduneword{htcondorce}{HTCondor-CE}{add def}  %anne added

\newduneabbrev{ccb}{CCB}{Computing Contributions Board}{The Computing Contributions is made up of institutional representatives for larger countried and laboratories. It meets annually to negotiate collaboration contributions to computing infrastructure.}

\newduneword{garfield}{Garfield}{A simulation program\cite{Veenhof:1998tt} for gaseous detectors developed at CERN.}

\newduneword{datatier}{data tier}{Differing data types produced in a processing sequence, for example, raw data, reconstructed, derived analysis sample, histograms.}

\newduneabbrev{gpu}{GPU}{Graphical Processing Unit}{Specialized computing hardware optimized for image processing.}

\newduneword{madx}{MAD-X}{framework that provides the de facto standard scripting language to describe particle accelerators, simulate beam dynamics and optimize beam optics at  \gls{cern}}

\newduneabbrev{cpu}{CPU}{Central Processing Unit}{A computing processor, when used as a unit of processing, it generally refers to a single core.}

\newduneword{grain}{GRAIN}{in \gls{sand}, a small cryostat containing liquid argon installed upstream of the straw-tube tracker inside the \gls{ecal}}

\newduneword{coldbox}{cold box}{Electronics and detectors are tested in cold boxes at CERN. These tests range from electronics tests at cryogenic temperature to full readout capable of reconstructing cosmic rays.f}
% coldbox is not needed; we have a def: \coldbox that renders as cold box. Anne

\newduneword{voms}{VOMS}{Virtual Organization Membership Service (in grid computing)}

\newduneword{jwt}{JWT}{\gls{json} web token}

\newduneword{ferry}{FERRY}{add def!!}

\newduneword{fifemon}{FIFEMON}{add def!!}

%\newduneword{enstore}{ENSTORE}{the mass storage system developed and implemented at \gls{fnal} as the primary data store for scientific data sets. It provides access to data on tape to/from a user's machine onsite over the local area network, or over the wide area network through the \gls{dcache} disk caching system}

\newduneword{perfsonar}{perfSONAR}{the performance Service-Oriented Network monitoring ARchitecture, a network measurement toolkit designed to provide federated coverage of paths and help to establish end-to-end usage expectations}

\newduneword{ipv6}{IPv6}{the most recent version of the Internet communications protocol that provides an identification and location system for computers on networks and routes traffic across the Internet}

\newduneword{monit}{MONIT}{\hideme{add def}}

\newduneword{garana}{GArAna}{provides a facility for making an analysis ntuple from information stored in \gls{garsoft} data products}

\newduneword{ci}{CI}{continuous integration}

\newduneword{datadis}{Data Dispatcher}{The Data Dispatcher communicates between running processing jobs and the data delivery systems.  It provides file location information and basic book-keeping on file access and transfers.}

\newduneabbrev{castor}{CASTOR}{CERN Advanced STORage manager}{CASTOR is a hierarchical storage system with tape and disk developed at CERN. It is being replaced by CTA.}

\newduneword{cta}{CTA}{CERN Tape Archive}{CTA is a hierarchical storage system with tape and disk developed at CERN. It is replacing CASTOR.}

\newduneabbrev{sonic}{SONIC}{Services for Optimized
Network Inference on Coprocessors}{Framework for implementing machine learning algorithms on coprocessors developed by Fermilab.}

\newduneword{jobsub}{JobSub}{JobSub\cite{box2014fife} is the Fermilab Intensity Frontier job submissions client that supports user submission of complex workflows to HT-Condor.}

\newduneword{hepcloud}{HEPCloud}{HEPCloud routes jobs to local or remote computing resources based on the policy for a particular experiment, workflow requirements, cost and efficiency of accessing the various resources. It expands the resources available to include \gls{hpc} centers and commercial cloud resources}

\newduneword{jobsub}{Jobsub}{a batch submission system to quickly get jobs running on the \gls{osg}}

\newduneword{wmsmon}{WMS monitoring}{add def}